|

### 2024-05-06 manuscript/paper.Rmd
Small changes and some notes in the results section.

---
title: |
  |  \Large Increased sexual dimorphism evolves in a fossil stickleback following ecological release from fish piscivores
author: | 
  | \large Allison Ozark$^{1,*}$, Matthew Stuart$^{2,3,*}$, Raheyma Siddiqui$^1$, Akhil Ghosh$^2$
  | \large Samantha Swank$^1$, Michael A. Bell$^4$, Gregory J. Matthews$^{2,3}$, and Yoel E. Stuart$^{1,+}$ 
  | \vspace{-1.1mm}
  | \large $^1$ Department of Biology, Loyola University Chicago, Chicago, IL, USA \vspace{-1.1mm}
  | \large $^2$ Department of Mathematics and Statistics, Loyola University Chicago, Chicago, IL, USA \vspace{-1.1mm}
  | \large $^3$ Center for Data Science and Consulting, Loyola University Chicago, Chicago, IL, USA \vspace{-1.1mm}
  | \large $^4$ University of California Museum of Paleontology, Berkeley, CA, USA \vspace{-1.1mm}
  | \large $^*$ Equal contribution \vspace{-1.1mm}
  | \large $^+$ Corresponding: ystuart@luc.edu \vspace{-1.1mm}
abstract: |
  | Everyone loves the stickle \vspace{2mm}
  | *Keywords*: Stickle
bibliography: references.bib
fontsize: 12pt
link-citations: true
linkcolor: cyan
urlcolor: cyan
output:
  pdf_document:
    df_print: kable
    number_sections: true
    keep_tex: true
header-includes:
 \usepackage{setspace}
 \setstretch{1.15}
 \usepackage{float}
 \floatplacement{figure}{t}
---

@@

| Trait Name           | Trait Code | Trait Description           |
|----------------------|------------|-----------------------------|
|Standard Length|	stl|	Distance from anterior tip of premaxilla to posterior end of last vertebra (hypural plate)|
|Dorsal Spine | mds |	Number of dorsal spines from 0 to 3|
|Dorsal Fin Ray|mdf|	Number of bones in the dorsal fin posterior to the third dorsal spine (i.e., soft dorsal fin rays)
|Anal Fin Ray |maf|	Number of bones in the anal fin posterior to the anal spine (i.e., soft dorsal fin rays)|
|Abdominal Vertebra|mav|	Number of vertebrae anterior to the first vertebra contacting an anal fin pterygiophore (Aguirre et al. 2014)|
|Caudal Vertebra|mcv	| Number of vertebrae posterior to and including the first vertebra contacting an anal fin pterygiophore (Aguirre et al. 2014)|
|Pterygiophore number|mpt|	Number of pterygiophores anterior to but excluding the pterygiophore under the third dorsal spine, which is immediately anterior to and contiguous with the dorsal fin|
|Pelvic Spine length|lps.sc	|Length from the base of one pelvic spine above its articulation with the pelvic girdle to its distal tip|
|Ectocoracoid|ect.sc|	Length between the anterior and posterior tips of the shoulder girdle base (i.e., ectocoracoid)|
|Pelvic Girdle|tpg.sc|	Length between the anterior to posterior tips along midline. If vestigial, the sum of longest anterior-posterior axis for the vestiges|
|Cleithrum length|cle.sc|	Length from free dorsal tip to ventral tip of the cleithrum on the anterior margin of the shoulder girdle (i.e., cleithrum)|
|Premaxilla|pmx.sc|	Length from the anterior tip of the premaxilla to the distal tip of the ascending process of the premaxilla |
|Dorsal Spine|Ds#.sc# = 1,2,or 3|	Length from the base of a dorsal spine above the pterygiophore to its distal tip along the anterior edge|
|Pterygiophore|lpt.sc|	Distance between the anterior to posterior tips of the pterygiophore immediately preceding the 3rd dorsal spine (when present)|

@@

We impute the missing gender for the fossil data by sampling from the posterior predictive distribution $P(\boldsymbol{W}_{fossil}|\boldsymbol{W}_{extant},\boldsymbol{Y})$ using the multiple imputation by chained equations (MICE) algorithm (@MICE) with predictive mean matching.  Traditionally, the choice for the number of completed dats sets is a relatively small number such as $M = 5$ or $M = 10$.  However, @ZhouReiter2009 recommend a larger number of imputed data sets if the data users intend on performing Bayesians analysis after imputation, which in this case, we do.  Therefore, the imputation algorithm is run to obtain a total of $M = 100$ completed datasets. In addition to this, @ZhouReiter2009 suggests rather than using Rubin's combining rules to combine across imputed data sets, instead pool all of the draws from the posterior distributions across all of the imputed data sets to estimate the posterior distributions of parameters of interest.  We proceed with our Bayesian analysis in this manner.

@@

<!-- Bayesian Analysis after multiple imputation @ZhouReiter2009: They recommend using a large number of imputations.  5 or 10 is too small.  We are using M = 100. -->

@@

@@

If $Y_{k,ij}$ is a discrete trait, the conventional method of modelling this data is by fitting a Poisson distribution. However, the Poisson distribution assumes that the mean and variance of $Y_{k,ij}$ are equal ($E(Y_{k,ij}) = \text{Var}(Y_{k,ij})$), while the empirical fossil data may have variances that are smaller than their respective means. To combat this issue, we propose to fit the discrete traits to a generalized Poisson model as defined in @GeneralizedPoisson. Specifically, if $X \sim GP(\lambda,\alpha)$, then 
\begin{align}
P(X = x) = \left\{\begin{array}{cc} \frac{(1 - \alpha)\lambda[(1 - \alpha)\lambda + \alpha x]^{x - 1} \exp\left\{-((1 - \alpha)\lambda  + \alpha x)\right\}}{x!} & (1 - \alpha)\lambda  + \alpha x \geq 0  \\ 0 & (1 - \alpha)\lambda  + \alpha x < 0 \end{array}\right.,
\label{eq:GP_pmf}
\end{align}
where $E(Y_{k,ij}) = \lambda$ and $Var(Y_{k,ij}) = \frac{\lambda}{(1 - \alpha)^2}$.  If $\alpha > 0$, then the variance is greater than the mean, called overdispersion; if $\alpha < 0$, then the variance is greater than the mean, called underdispersion; if $\alpha = 0$, then the model degenerates to a Poisson distribution.

@@

Bayesian Analysis after multiple imputation @ZhouReiter2009: They recommend using a large number of imputations.  5 or 10 is too small.  We are using M = 100.

# Results {#sec:results}

>>>>

---
title: |
  |  \Large Increased sexual dimorphism evolves in a fossil stickleback following ecological release from fish piscivores
author: | 
  | \large Allison Ozark$^{1,*}$, Matthew Stuart$^{2,3,*}$, Raheyma Siddiqui$^1$, Akhil Ghosh$^{2,3}$
  | \large Samantha Swank$^1$, Michael A. Bell$^4$, Gregory J. Matthews$^{2,3}$, and Yoel E. Stuart$^{1,+}$ 
  | \vspace{-1.1mm}
  | \large $^1$ Department of Biology, Loyola University Chicago, Chicago, IL, USA \vspace{-1.1mm}
  | \large $^2$ Department of Mathematics and Statistics, Loyola University Chicago, Chicago, IL, USA \vspace{-1.1mm}
  | \large $^3$ Center for Data Science and Consulting, Loyola University Chicago, Chicago, IL, USA \vspace{-1.1mm}
  | \large $^4$ University of California Museum of Paleontology, Berkeley, CA, USA \vspace{-1.1mm}
  | \large $^*$ Equal contribution \vspace{-1.1mm}
  | \large $^+$ Corresponding: ystuart@luc.edu \vspace{-1.1mm}
abstract: |
  | Everyone loves the stickle \vspace{2mm}
  | *Keywords*: Stickle
bibliography: references.bib
fontsize: 12pt
link-citations: true
linkcolor: cyan
urlcolor: cyan
output:
  pdf_document:
    df_print: kable
    number_sections: true
    keep_tex: true
header-includes:
 \usepackage{setspace}
 \setstretch{1.15}
 \usepackage{float}
 \floatplacement{figure}{t}
---

@@

| Trait Name           | Trait Code | Trait Description           |
|----------------------|------------|-----------------------------|
|Standard Length|	`stl` |	Distance from anterior tip of premaxilla to posterior end of last vertebra (hypural plate)|
|Dorsal Spine | `mds` |	Number of dorsal spines from 0 to 3|
|Dorsal Fin Ray|`mdf`|	Number of bones in the dorsal fin posterior to the third dorsal spine (i.e., soft dorsal fin rays)
|Anal Fin Ray |`maf`|	Number of bones in the anal fin posterior to the anal spine (i.e., soft dorsal fin rays)|
|Abdominal Vertebra|mav|	Number of vertebrae anterior to the first vertebra contacting an anal fin pterygiophore (Aguirre et al. 2014)|
|Caudal Vertebra|mcv	| Number of vertebrae posterior to and including the first vertebra contacting an anal fin pterygiophore (Aguirre et al. 2014)|
|Pterygiophore number|mpt|	Number of pterygiophores anterior to but excluding the pterygiophore under the third dorsal spine, which is immediately anterior to and contiguous with the dorsal fin|
|Pelvic Spine length|lps.sc	|Length from the base of one pelvic spine above its articulation with the pelvic girdle to its distal tip|
|Ectocoracoid|ect.sc|	Length between the anterior and posterior tips of the shoulder girdle base (i.e., ectocoracoid)|
|Pelvic Girdle|tpg.sc|	Length between the anterior to posterior tips along midline. If vestigial, the sum of longest anterior-posterior axis for the vestiges|
|Cleithrum length|cle.sc|	Length from free dorsal tip to ventral tip of the cleithrum on the anterior margin of the shoulder girdle (i.e., cleithrum)|
|Premaxilla|pmx.sc|	Length from the anterior tip of the premaxilla to the distal tip of the ascending process of the premaxilla |
|Dorsal Spine|Ds#.sc# = 1,2,or 3|	Length from the base of a dorsal spine above the pterygiophore to its distal tip along the anterior edge|
|Pterygiophore|lpt.sc|	Distance between the anterior to posterior tips of the pterygiophore immediately preceding the 3rd dorsal spine (when present)|

@@

We impute the missing gender for the fossil data by sampling from the posterior predictive distribution $P(\boldsymbol{W}_{fossil}|\boldsymbol{W}_{extant},\boldsymbol{Y})$ using the multiple imputation by chained equations (MICE) algorithm (@MICE) with predictive mean matching.  Traditionally, the choice for the number of completed dats sets is a relatively small number such as $M = 5$ or $M = 10$.  However, @ZhouReiter2010 recommend a larger number of imputed data sets if the data users intend on performing Bayesians analysis after imputation, which in this case, we do.  Therefore, the imputation algorithm is run to obtain a total of $M = 100$ completed datasets. In addition to this, @ZhouReiter2010 suggests rather than using Rubin's combining rules to combine across imputed data sets, instead pool all of the draws from the posterior distributions across all of the imputed data sets to estimate the posterior distributions of parameters of interest.  We proceed with our Bayesian analysis in this manner.

@@

<!-- Bayesian Analysis after multiple imputation @ZhouReiter2010: They recommend using a large number of imputations.  5 or 10 is too small.  We are using M = 100. -->

@@

\textcolor{red}{I think we need to move a lot of these details to the appendix.  }

@@

If $Y_{k,ij}$ is a discrete trait, the conventional method of modelling this data is by fitting a Poisson distribution. However, the Poisson distribution assumes that the mean and variance of $Y_{k,ij}$ are equal ($E(Y_{k,ij}) = \text{Var}(Y_{k,ij})$), while the empirical fossil data may have variances that are smaller than their respective means. To combat this issue, we propose to fit the discrete traits to a generalized Poisson model as defined in @GeneralizedPoisson. Specifically, if $X \sim GP(\lambda,\alpha)$, then 
\begin{align}
P(X = x) = \left\{\begin{array}{cc} \frac{(1 - \alpha)\lambda[(1 - \alpha)\lambda + \alpha x]^{x - 1} \exp\left\{-((1 - \alpha)\lambda  + \alpha x)\right\}}{x!} & (1 - \alpha)\lambda  + \alpha x \geq 0  \\ 0 & (1 - \alpha)\lambda  + \alpha x < 0 \end{array}\right.,
\label{eq:GP_pmf}
\end{align}
where $E(X) = \lambda$ and $Var(X) = \frac{\lambda}{(1 - \alpha)^2}$.  If $\alpha > 0$, then the variance is greater than the mean, called overdispersion; if $\alpha < 0$, then the variance is greater than the mean, called underdispersion; if $\alpha = 0$, then the model degenerates to a Poisson distribution.

@@

# Results {#sec:results}
Greg notes: 
- mdf and maf have interesting trends.  
- ds3 flips over time.  Females have larger values to start and over the course of the time frame wer are looking at males end up with larger mean values.  
- Most of the other traits have mean differences that are relatively constant over time.
